Multitasking Interfaces on Touchscreen Mobile Devices

### 2015-12-10 sigchi-latex-proceedings 2/final-paper.tex
added sections except results

The mechanism by which users switch active apps varies between devices. Lists of applications, search methods, or gestures are just some of the interactions to switch apps on smartphones. The most common way to switch apps is to open the application switching interface, or rather, the multitasking interface which varies by device. In the two major operating systems for smartphones, Android and iOS, each application is displayed in a list where the user can scroll through to select the desired application.

@@

Switching between apps incurs a cost in terms of task completion time. Nagata [5] examined the impact of digital interruptions on task performance on mobile and desktop devices, and Leiva et al. [3] studied the cognitive costs associated with switching apps. An interruption, say in the form of an app notification, can cause a user to multitask or switch apps by shifting attention from the task to the interruption [5]. Both studies found that task completion time increased when users switched between apps; in Leiva et al.'s study, that completion time was increased by a factor of four [3]. It was found that mobile web tasks with interruptions take longer to complete than on desktop, but when the interruption is expected, performance time decreased on a mobile device [5]. As a result, if a user is expecting to switch apps, they can complete tasks between apps more efficiently.

Given the cost to task completion time from switching apps, there have been many studies done on how to increase task resumption and completion time when multitasking. Leiva [2] explored different techniques to improve task resumption time in a multi-tab browser environment, and Lottridge et al. [4] investigated how the design of applications encourages or discourages multitasking behavior in general. Leiva investigated the effects of a tool designed to remind users of their position on the screen prior to app switching as well as the last interacted element and cursor position, while Lottridge et al. looked at making users more aware of the time spent in "work" versus "non-work" categories of websites. Lottridge et al. found that by making users more aware of the passage of time, users had fewer webpage tabs open, fewer tab switches and shorter sessions within webpages [4], indicating that users were more efficient with their time spent on the web. Leiva found that the use of his tool improved task resumption time and task completion time.

Card and Henderson [1] observed that users interacting with an interface often organize tasks along a number of different dimensions, the most important of them being "locality" of tasks. The study found that users often form clusters of windows corresponding to a specific task. This insight led to development of interfaces that produce "virtual views" that allow users to focus their interactions within semantically meaningful clusters of windows. Modern smartphone multitasking interfaces sort opened apps in chronological order, but if a user is switching between two or three apps a time to perform a task, those apps in smartphone multitasking interfaces will be sorted together at the front of the list. If the apps are "clustered" to match with the user's mental model, this could result in faster and less error-prone application switching. In fact, Oliver et al. investigated a tool that indicated which windows in a multitasking interface were semantically related [6]. The use of this tool resulted in significantly more efficient task completion time, but it was noted that there was a negative effect when windows in the multitasking interface were reordered if the user interrupted the main task to switch to a different task [6]. This is what currently occurs in smartphone multitasking interfaces, which is why we intend to investigate the speed and accuracy of modern smartphone multitasking interfaces.

@@

H1: Users will be faster at switching between applications in the stacked interface.\\
H2: Users will have greater accuracy when switching between applications in the non-stacked interface. \\

@@

\section{Methodology}
\subsection{Apparatus}
Two LG Nexus 5 phones were used as the apparatus for the experiment, one loaded with the stacked multitasking interface and the other with the non stacked interface. The stacked interface uses Android OS version 5.1 (Lollipop) while the non-stacked uses Android OS 4.4.4 (KitKat). Both Android OS versions used were official Google releases and did not include any carrier or manufacturer modifications.

The Nexus 5 was released two years ago and has a 4.95 inch screen with a Snapdragon 800 Quad-Core processing chip. These specifications put it in the mid-range of current Android phones. We felt that using the Nexus 5 gives us a more accurate representation of what an average Android user would be using in the real world. This will reduce the bias of unfamiliarity and novelty when testing.

@@

\subsection{Participants}
Sixteen participants (ages 19-26) were recruited and did not receive monetary compensation for their participation. They all were currently pursuing an undergraduate degree at the University of Toronto. All participants met the criteria of previous experience with the Android operating system and no visual or hand dexterity disabilities. Participants were randomly assigned to begin with the stacked interface (n=8) or non-stacked interface (n=8). The study was carried out in a meeting room at the Mobile Application Development Lab (MADLab) which is located at the University of Toronto.

\subsection{Experimental Design}
The experiment had a one-way within-subject design with the multitasking interface (stacked and non-stacked) acting as a two-level independent variable. Order effects were counterbalanced by switching the order of testing with two groups of participants. The first group tested the stacked interface first on the Lollipop phone, while the second group tested the non-stacked interface first on the KitKat phone. Time to completion and accuracy served as the dependent variables. Participants performed the set of tasks twice for each stacked and non-stacked multitasking interfaces with two different experimenters.

Randomizing participants into groups allowed us to minimize the variance any participants may have had in terms of greater experience or comfort with one interface over the other. Since the experimental tasks are repetitive, the participants were susceptible to the learning-effect. However, counterbalancing based on the two levels of the independent variable allowed us to account for it.

@@

Each task was designed to cover one of three common interactions. First was a minimal interaction, where the application switches was limited to one or two switches between three open applications. Second was a longer task where the participant switched between three or more applications. In this task, participants had to scroll through the multitasking interface, which presented a non-ordered list of apps to find the app they were to switch to. Last was a repetitive switching task, a common scenario where the participant repeatedly switched between two apps to achieve a specified goal. Combined with a within-subject design where the participant performs these tasks multiple times on each level gave us a comprehensive set of data points to test the hypotheses.

Each participant executed each block twice, once on each multi-tasking interface (stacked and non-stacked). There were six trials in total per participant. Each participant was handed a sheet outlining all the tasks for each block. Each block of tasks was preceded by a short break for the experimenter to setup the device and demonstrate the tasks, if requested by the participant. Participants were instructed to begin tasks and were observed for any errors. Errors were recorded as a total per task block. Tasks were completed at the participant's? own pace and errors (if any) were corrected at their discretion. Experimenter logged the data for each block of tasks on each interface. At the commencement of the experiment, the participant completed an exit survey. The whole experiment took about 20 minutes to complete per participant.

\subsection{Measures}
The dependent variables in this study were time spent in the multitasking interface, which corresponds to time taken to switch between apps, and the  accuracy when switching apps. Time to completion is measured as when the participant opens the multitasking interface to when the participant opens the target app. Accuracy is defined as when a participant does not open the target app in the task and instead either opens a different app or goes to the home screen. Measuring time to completion and accuracy will show which interface allows for fastest and least error-prone application switching.

@@

\section{Discussion}
While the trends in the speed and accuracy measurements of our participants was statistically insignificant between the two interfaces, a discussion can still take place. The stacked interface was perceived by the participants to be faster, which held when examining the data. This is inline with previous research, such as the work of Warr and Chi [7]. However, a follow-up question is presented: why is stacked interface perceived as quicker? It?s possible this is due to the fact that the motion of the cards is toward the user rather than down. In addition, with the stacked interface the user can control the sensitivity of the scroll by scrolling from higher up on the screen. This allows users to more quickly traverse a long list of apps at the cost of some granularity of motion.

\section{Conclusion}
Overall, the experiment showed that the differences between the two interfaces are statistically insignificant for accuracy and speed. The stacked  interface was rated as generally more pleasant and quicker, which could be a reason Google updated Android?s default multitasking interface.

One limitation of our participant group was their familiarity with the two operating systems. From the starting survey 9 participants were mostly familiar with Lollipop whereas only 3 were mostly familiar with KitKat. Another limitation is that the study only tested one model of phone with stock Android. The real consumer landscape for Android contains many different sized phones running various modified versions of Android.

A stacked interface may only provide nominal speed and accuracy increases when compared to a non-stacked interface. As a result, future research should focus on alternate means to facilitate app switching as opposed to the conventional chronological list of applications. Throughout all mobile multitasking interfaces, almost exclusively, the "most recently used" rule is used to display applications in the list. Future research could look into using different sorting schemes to predict which application a user will need next. Perhaps machine learning could be applied to predict what application may be used next as is being tested in Apple?s new recents menu.

>>>>

The mechanism by which users switch active apps varies between devices. Lists of applications, search methods, or gestures are common interactions to switch apps on smartphones. The most common way to switch apps is to open the application switching interface, or rather, the multitasking interface which varies by device. In the two major operating systems for smartphones, Android and iOS, each application is displayed in a list where the user can scroll through to select the desired application.

@@

Switching between apps incurs a cost in terms of task completion time. Nagata [5] examined the impact of digital interruptions on task performance on mobile and desktop devices, and Leiva et al. [3] studied the cognitive costs associated with switching apps. An interruption, say in the form of an app notification, can cause a user to multitask or switch apps by shifting attention from the task to the interruption [5]. Both studies found that task completion time increased when users switched between apps; in Leiva et al.?s study, that completion time was increased by a factor of four [3]. It was found that mobile web tasks with interruptions take longer to complete than on desktop, but when the interruption is expected, performance time decreased on a mobile device [5]. As a result, if a user is expecting to switch apps, they can complete tasks between apps more efficiently.

Given the cost to task completion time from switching apps, there have been many studies done on how to increase task resumption and completion time when multitasking. Leiva explored different techniques to improve task resumption time in a multi-tab browser environment [2], and Lottridge et al. investigated how the design of applications encourages or discourages multitasking behavior in general [4]. Leiva investigated the effects of a tool designed to remind users of their position on the screen prior to app switching as well as the last interacted element and cursor position, while Lottridge et al. looked at making users more aware of the time spent in ?work? versus ?non-work? categories of websites. Lottridge et al. found that by making users more aware of the passage of time, users had fewer webpage tabs open, fewer tab switches and shorter sessions within webpages [4], indicating that users were more efficient with their time spent on the web. Leiva found that the use of his tool improved task resumption time and task completion time.

Card and Henderson [1] observed that users interacting with an interface often organize tasks along a number of different dimensions, the most important of them being "locality" of tasks. The study found that users often form clusters of windows (apps) corresponding to a specific task. This insight led to development of interfaces that produce "virtual views" that allow users to focus their interactions within semantically meaningful clusters of windows. Modern smartphone multitasking interfaces sort opened apps in temporal order, but if a user is switching between two or three apps a time to perform a task, those apps in smartphone multitasking interfaces will be sorted together at the front of the list. If the apps are ?clustered? to match with the user?s mental model, this could result in faster and less error-prone application switching. In fact, Oliver et al. investigated a tool that indicated which windows in a multitasking interface were semantically related [6]. The use of this tool resulted in significantly more efficient task completion time, but it was noted that there was a negative effect when windows in the multitasking interface were reordered if the user interrupted the main task to switch to a different task [6]. This is what currently occurs in smartphone multitasking interfaces, which is why we intend to investigate the speed and accuracy of modern smartphone multitasking interfaces.

@@

H1: Users will be faster at switching between applications in the stacked interface. \\
H2: Users will have greater accuracy when switching between applications in the non-stacked interface. \\

@@

\section{Methodology}
\subsection{Apparatus}
Two LG Nexus 5 phones were used as the apparatus for the experiment, one loaded with the stacked multitasking interface and the other with the non stacked interface. The stacked interface was on Android OS version 5.1 (Lollipop) while the non-stacked was on Android OS 4.4 (KitKat). Both Android OS versions used were official Google releases and did not include any carrier or manufacturer modifications.

The Nexus 5 was released in October 2013. It has a 4.95 inch screen, 2GB of RAM and is powered by a 2.26GHz Snapdragon 800 Quad-Core processing chip. These specifications put it in the mid-range of current Android phones. Using the Nexus 5 in the experiment provides an accurate representation of what an average Android user would be using in the real world. This reduces the bias of unfamiliarity and novelty when testing.

@@

\subsection{Participants}
Sixteen participants between the ages of 19 and 29 (mean age = 22.5 yrs, females = 7, males = 9) were recruited and did not receive monetary compensation for their participation. They all were currently pursuing an undergraduate degree at the University of Toronto. All participants met the criteria of previous experience with the Android operating system and no visual or hand dexterity disabilities. Participants were randomly assigned to begin the experiment with the stacked interface (n=8) or non-stacked interface (n=8). The study was carried out in a meeting room at the Mobile Application Development Lab which is located at the University of Toronto.

\subsection{Experimental Design}
The experiment had a one-way within-subject design with the multitasking interface (stacked and non-stacked) acting as a two-level independent variable. Order effects were counterbalanced by switching the order of testing with two groups of participants. The first group tested the stacked interface first on the Lollipop phone, while the second group tested the non-stacked interface first on the KitKat phone. Task completion time and accuracy were the dependent variables. Participants performed the set of tasks twice for each stacked and non-stacked multitasking interfaces with two different experimenters.

Randomizing participants into groups minimized the variance any participants may have had in terms of greater experience or comfort with one interface over the other. Since the experimental tasks are repetitive, the participants were susceptible to the learning-effect. However, counterbalancing based on the two levels of the independent variable allowed us to account for it.

@@

Each task was designed to cover one of three common interactions. First was a minimal interaction, where the application switches were limited to one or two switches between three open applications. Second was a longer task where the participant switched between three or more applications. In this task, participants had to scroll through the multitasking interface, which presented a non-ordered list of apps to find the app they were to switch to. Last was a repetitive switching task, a common real-life scenario where the participant repeatedly switched between two apps to achieve a specified goal. Combined with a within-subject design where the participant performs these tasks multiple times on each level gave us a comprehensive set of data points to test the hypotheses.

Each participant executed each block twice, once on each multi-tasking interface. There were six trials in total per participant. Each participant was handed a sheet outlining all the tasks for each block. Each block of tasks was preceded by a short break for the experimenter to setup the device and demonstrate the tasks, if requested by the participant. Participants were instructed to begin tasks and were observed for any errors. Errors were recorded as a total per task block. Tasks were completed at the participant's? own pace and errors (if any) were corrected at their discretion. Experimenter logged the data for each block of tasks on each interface. At the commencement of the experiment, the participant completed an exit survey. The whole experiment took about 20 minutes to complete per participant.

\subsection{Measures}
The dependent variables in this study were time spent in the multitasking interface, which corresponds to time taken to switch between apps, and the  accuracy when switching apps. Task completion time is measured as when the participant opens the multitasking interface to when the participant opens the target app. Accuracy is defined as when a participant does not open the target app in the task and instead either opens a different app or goes to the home screen. Measuring task completion time and accuracy will show which interface allows for fastest and least error-prone application switching.

@@

\section{Discussion}
Though the results were not statistically significant, our data did show a slight trend that the stacked interface was, on average, quicker and resulted in less errors when performing the tasks. For H1, the differences in task completion time between the two interfaces supports our hypothesis that the stacked interface is quicker. As a result, this corroborates the work done by Warr and Chi [7]. H2 was not supported by our data, as the stacked interface, on average, had less errors while performing the tasks than the non-stacked interface. There was huge variation in participant accuracy and the difference was not close to being statistically significant. This could be due to the participants? level of familiarity with Android instead of a result of the interface design. 
	The two multitasking interfaces required the same number of input actions: one tap to enter the interface and one tap to exit. The primary differences between the interfaces was the amount of scrolling that was required to reach an app, as well as the target area to select an application. The stacked interface overlaps application screens; as a result, the list is compressed and a user can view more apps at once. This comes at the cost of the size of the touchable area that is available to select an application. Furthermore, on the stacked interface scrolling from higher up on the list increases the speed at which the list is scrolled; the scroll speed is always directly proportional to the speed of the user?s finger on the non-stacked interface. Though the stacked interface was, on average, slightly quicker we hypothesize that this effect is due to the perceived motion of scrolling through the list to be more pronounced in the stacked interface. 
	The stacked interface was preferred by the majority of participants. The perceived speed and interface preference were highly correlated; 100\% of participants who rated the stacked interface as quicker stated it was their preferred interface. Thus, perceived speed is important to interface preference.

\section{Conclusion}
Overall, the experiment showed that the differences between the two interfaces are statistically insignificant for accuracy and speed. The stacked interface was rated as generally more pleasant and quicker, which could be a reason why Google updated Android?s default multitasking interface (non-stacked) to the current stacked interface.
	One limitation of our participant group was their familiarity with the two operating systems. From the starting survey 9 participants were mostly familiar with Lollipop whereas only 3 were mostly familiar with KitKat. Another limitation could be due to the small sample size of participants and ultimately the number of switches that were performed. More trials may need to be performed to notice a statistically significant difference.
	A stacked interface may only provide nominal speed and accuracy increases when compared to a non-stacked interface. As a result, future research should focus on alternate means to facilitate app switching as opposed to the conventional chronological list of applications. Throughout all mobile multitasking interfaces, almost exclusively, the ?most recently used? rule is used to display applications in the list. Future research could look into using different sorting schemes to predict which application a user will need next.
